Contracts for unions and intersections are hard for Nickel

### 2021-05-24 paper.tex
Review mode, preparing for submission
Typo
Smallfixes
sec 2 small changes

\documentclass[sigplan,10pt]{acmart}
%\usepackage{amssymb}
\usepackage{mathtools,amsmath,amsfonts,amsxtra,amsthm,mathrsfs,mathpartir}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{array}
\usepackage[toc,page]{appendix}
\usepackage{graphicx}
\usepackage{varioref}
\usepackage{verbatim,comment}
\usepackage{longtable}
\usepackage{tikz}
\usepackage{url}
\usepackage{hyperref}

@@

For instance, we could define a flat contract \nickel{even} as
\nickel{fun x => x \% 2 == 0}, that checks
whether a number is even by checking the remainder of dividing it by 2.
And then use it to check that a function always returns an
even number, when applied to an \nickel{Integer}, as shown in
Figure \ref{fig:num-to-even}.
Or even, we could come up with a contract that checks that a function
maps \nickel{0} to \nickel{0}, as can be seen on Figure
\ref{fig:0-to-0}.

@@

\begin{figure}[h]
  \begin{center}
\begin{lstlisting}[language=nickel]
fun f => f 0 == 0
\end{lstlisting}
    \end{center}
  \caption{A contract that checks that a function maps 0 to 0}
  \label{fig:0-to-0}
\end{figure}
\unsure{Do we need deeper examples?
Maybe something showing that flat contracts could evaluate
a function.}

This kind of contracts are present on many different languages,
for instance, the Eiffel programming language, the precursor
of the Design by Contract ideals, allows to assert
these kinds of expression as pre- and post-conditions on
functions\cite{EiffelDesignByContract}.
A somewhat newer language, the D programming
language, allows, among other things, to have invariants on classes,
written with flat contracts, that are
dinamically checked at the frontiers of every public method \cite{DLangContracts}.

Finally, the Racket programming language also has a system to work with
contracts that allows to define flat contracts
\footnote{And a gradual typing system
built on top of that, but that goes beyond the scope of this
work.}
, and
to compose them with other kinds of dynamic checks,
like higher order contracts or a lightweight take on union
and intersection contracts\cite{RacketContracts}.
\todo{Talk about this, probably later on the paper.
See Related Works section}

Even if flat contracts may seem like too permissive, the ability to make
non trivial checks on values (that is, to validate their output) is expected
by programmers on dynamic languages, where testing and runtime
assertions are extremely useful tools for safe programming, and
programmers are used to express complex predicates
over values.

@@

One crucial optimization is
inlining(Figure~\ref{fig:optimizations-inlining-ex}). Functional programs tend
to make heavy use of functions, and a function call is a costly operation from
the point of view of low-level execution. Inlining eliminates a function
application by directly substituting the function for its definition at compile
time (or before execution, for interpreted language). This is especially
efficient for small functions that are called repeatedly.

\begin{figure}
  \begin{center}
\begin{lstlisting}[language=nickel,title={Source program}]
let elem = fun elt =>
  lists.any (fun x => x == elt) in

@@

While inlining expands an expression by substituting a definition for its value,
an opposite transformation is sometimes beneficial, for example when a composite
expression is repeated several times. This causes the same expression to be
wastefully recomputed at each occurrence. Common subexpression elimination (CSE)
consist in introducing a new identifier for this expression and using the
identifier in place of the original occurrences
(Figure~\ref{fig:optimizations-cse-ex}), computing the result once and for all.

\begin{figure}
  \begin{center}
\begin{lstlisting}[language=nickel,title={Source program}]
let elemAtOrLast = fun index list =>
  if index > lists.length list - 1 then
    lists.elemAt (lists.length list - 1) list
  else
    lists.elemAt index list
\end{lstlisting}
\begin{lstlisting}[language=nickel,title={Optimized program}]
let elemAtOrLast = fun index list =>
  let l = lists.length list - 1 in
  if index > l then
    lists.elemAt l list
  else
    lists.elemAt index list
\end{lstlisting}
  \end{center}
\caption{Common subexpression elimination}
\label{fig:optimizations-cse-ex}
\end{figure}

@@

\subsection*{Unions}
Unions are fundamental building blocks of programs. A union holds a value whose
shape is known to be in a fixed set of alternatives: for example, a value that
can be either a boolean or a string. They correspond logically to conjunctions.
\unsure{disjunctions, right?}
Used in combination with products, they can represent arbitrary tree-like
structures, allowing to model a large class of data
(\ref{fig:union-adt-ex}).

@@

\subsubsection*{Union contracts are hard to brake}

@@

Altough this is true, it's a weaker statement of what we'd like to have,
since it rules out continuations like $(\lambda f.~((f~3)~+~4,~not~(f ~ true)))$,
where $(\square 3)~+~4 \in \llbracket Number \rightarrow Number \rrbracket^-$
but $(\square 3)~+~4 \not \in \llbracket Boolean \rightarrow Boolean \rrbracket^-$,
and dually for $not~(f ~ true)$.
Which would mean, according to their semantics, that
the continuation presented above would not
be in $\llbracket Number \rightarrow Number \cap Boolean \rightarrow Boolean \rrbracket^-$,
even though our intuition (and the
introduction to the paper) indicates that it should.

@@

@@

Unions are fundamental and ubiquitous building blocks of program data. A union
only holds one value whose shape is only known to be in a fixed set of
alternatives: for example, either a boolean or a string. When used in combination
with products, they can represent arbitrary tree-like structures, allowing to
model a large class of data.

>>>>

\documentclass[sigplan,10pt,review]{acmart}
%\usepackage{amssymb}
\usepackage{mathtools,amsmath,amsfonts,amsxtra,amsthm,mathrsfs,mathpartir}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{array}
\usepackage[toc,page]{appendix}
\usepackage{graphicx}
\usepackage{varioref}
\usepackage{verbatim,comment}
\usepackage{longtable}
\usepackage{tikz}
\usepackage{url}
\usepackage{hyperref}

@@

For instance, we could define a flat contract \nickel{even} as
\nickel{fun x => x \% 2 == 0}, that checks
whether a number is even by checking the remainder of dividing it by 2.
And then use it to check that a function always returns an
even number, when applied to an \nickel{Integer}, as shown in
Figure \ref{fig:num-to-even}.
Or even, we could come up with a contract that checks that a function
maps \nickel{true} to \nickel{false} and \nickel{false} to \nickel{true},
and use it as a contract for the \nickel{not} function, as can be seen on Figure
\ref{fig:checked-not}.

@@

\begin{figure}[h]
  \begin{center}
\begin{lstlisting}[language=nickel]
let notC = fun f => (f true == false) and (f false == true) in
let checkedNot = not @ notC
\end{lstlisting}
\end{center}
\caption{A contract that checks the not function behaves as expected}
\label{fig:checked-not}
\end{figure}

This type of contracts are present on many different languages,
for instance, the Eiffel programming language, the precursor
of the Design by Contract ideals, allows to assert
these kinds of expression as pre- and post-conditions on
functions\cite{EiffelDesignByContract}.
A somewhat newer language, the D programming
language, allows, among other things, to have invariants on classes,
written with flat contracts, that are
dinamically checked at the frontiers of every public method \cite{DLangContracts}.

Finally, the Racket programming language also has a system to work with
contracts, powerful enough to define flat contracts
\footnote{And a gradual typing system
built on top of that, but that goes beyond the scope of this
work.}
, and
to compose them with other kinds of dynamic checks,
like higher order contracts or a lightweight take on union
and intersection contracts\cite{RacketContracts}.
\todo{Talk about this, probably later on the paper.
See Related Works section}

Even if flat contracts may seem like too permissive, the ability to make
non trivial checks on values (that is, to validate their output) is expected
by programmers on dynamic languages, where testing and runtime
assertions are extremely useful tools for safe programming, and
programmers are accustomed to express complex predicates
over values.

@@

One crucial optimization is
inlining (Figure \ref{fig:optimizations-inlining-ex}). Functional programs tend
to make heavy use of functions, and a function call is a costly operation from
the point of view of low-level execution. Inlining eliminates a function
application by directly substituting the function for its definition at compile
time (or before execution, for interpreted language). This is especially
efficient for small functions that are called repeatedly.

\info{I think it's better to use [h] on figures, we can discuss it}
\begin{figure}[h]
  \begin{center}
\begin{lstlisting}[language=nickel,title={Source program}]
let elem = fun elt =>
  lists.any (fun x => x == elt) in

@@

While inlining expands an expression by substituting a definition for its value,
an opposite transformation is sometimes beneficial, for example when a composite
expression is repeated several times. This causes the same expression to be
wastefully recomputed at each occurrence. Common subexpression elimination (CSE)
consists in introducing a new identifier for this expression and using the
identifier in place of the original occurrences
(Figure \ref{fig:optimizations-cse-ex}), computing the result once and for all.

\begin{figure}[h]
  \begin{center}
\begin{lstlisting}[language=nickel,title={Source program}]
let elemAtOrLast = fun index list =>
  if index > lists.length list - 1 then
    lists.elemAt (lists.length list - 1) list
  else
    lists.elemAt index list
\end{lstlisting}
\begin{lstlisting}[language=nickel,title={Optimized program}]
let elemAtOrLast = fun index list =>
  let l = lists.length list - 1 in
  if index > l then
    lists.elemAt l list
  else
    lists.elemAt index list
\end{lstlisting}
  \end{center}
\caption{Common subexpression elimination}
\label{fig:optimizations-cse-ex}
\end{figure}

@@

\subsection*{Unions}
Unions are fundamental building blocks of programs. A union holds a value whose
shape is known to be in a fixed set of alternatives: for example, a value that
can be either a boolean or a string. They correspond logically to disjunctions.
\unsure{disjunctions, right?}
Used in combination with products, they can represent arbitrary tree-like
structures, allowing to model a large class of data
(\ref{fig:union-adt-ex}).

@@

\subsubsection*{Union contracts are hard to break}

@@

Altough this is true, it's a weaker statement of what we'd like to have,
since it rules out continuations like $(\lambda f.~((f~3)~+~4,~not~(f ~ true)))$,
where $(\square 3)~+~4 \in \llbracket Number \rightarrow Number \rrbracket^-$
but $(\square 3)~+~4 \not \in \llbracket Boolean \rightarrow Boolean \rrbracket^-$,
and dually for $not~(f ~ true)$.
Which would mean, according to their semantics, that
the continuation presented above would not
be in $\llbracket Number \rightarrow Number \cap Boolean \rightarrow Boolean \rrbracket^-$,
even though our intuition (and the
introduction to the paper) indicates that it should.
\todo{Simplify example}

@@

\todo{Remove parts that are already on the main part}

@@

\todo{Specify: union types? more general concept?}{Unions} are fundamental and ubiquitous building blocks of program data. A union
only holds one value whose shape is only known to be in a fixed set of
alternatives: for example, either a boolean or a string. When used in combination
with products, they can represent arbitrary tree-like structures, allowing to
model a large class of data.
